---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# AssignParents

<!-- badges: start -->
<!-- Add R-CMD-check badge after enabling GitHub Actions -->
<!-- Add license badge after adding MIT license -->
[![R-CMD-check](https://github.com/pjoppenh/AssignParents/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pjoppenh/AssignParents/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**AssignParents** is a post-denoising refinement tool for ASV-based metabarcoding workflows.  It is meant to address the issue of Phred-score bias inherent to metabarcoding data and was originally conceptualized in:
Oppenheimer et al., *ISME Communications* (2024).
https://academic.oup.com/ismecommun/article/5/1/ycaf124/8209421.
This package builds upon the framework outlined in the paper by reconciling high-confidence ("CLEAN") and permissive ("DIRTY") ASV tables using:

- Sequence identity filtering
- Prevalence-aware presence similarity
- Abundance similarity
- Dominance filtering
- Deterministic parent selection

AssignParents is designed to reduce residual artifact ASVs and improve quantitative stability in downstream ecological and statistical analyses.
 
---

## Installation

Install the development version from GitHub:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("pjoppenh/AssignParents")
```

---

## Inputs and Outputs

`assign_parents()` requires two ASV count matrices:

- `clean_tab`: a high-confidence reference set of real sequence variants
- `dirty_tab`: a more permissive set of sequence variants with negligible read filtering.

Both tables must:

- Have ASVs as rows
- Have samples as columns
- Contain numeric count data
- Use rownames that uniquely identify each ASV

The corresponding sequence inputs:

- `clean_seqs`
- `dirty_seqs`

must be named character vectors of nucleotide sequences where:

- Names match the rownames of the respective count matrix
- Vector length equals the number of rows in the corresponding matrix

### Returns

The function returns a list containing:

- `collapsed_table`: CLEAN table with assigned DIRTY counts merged
- `mapping`: DIRTY → CLEAN parent assignments
- Identity, presence, abundance, and dominance score matrices
- Optional unassigned ASV tables (if `keep_unassigned = TRUE`)

All operations preserve total counts exactly.

---

## Example

```{r example}
library(AssignParents)

clean_seqs <- c(
  C1 = "ACGTACGTACGT",
  C2 = "GGGGACGTACGT"
)

dirty_seqs <- c(
  D1 = "ACGTACGTACGT",
  D2 = "ACGTACGTACGA"
)

clean_tab <- matrix(
  c(5,1,
    2,3),
  nrow = 2,
  dimnames = list(names(clean_seqs), c("S1","S2"))
)

dirty_tab <- matrix(
  c(1,0,
    0,1),
  nrow = 2,
  dimnames = list(names(dirty_seqs), c("S1","S2"))
)

out <- assign_parents(
  clean_tab,
  dirty_tab,
  clean_seqs,
  dirty_seqs,
  min_identity = 0.95,
  presence_cutoff = 1,
  min_n = 1,
  keep_unassigned = TRUE
)

out$collapsed_table
```

---

## Key Parameters

Only the most commonly adjusted parameters are listed here.  
See `?assign_parents` for full documentation.

- `min_identity`
  Minimum percent identity required for a DIRTY ASV to be considered for a CLEAN parent.

- `presence_cutoff`
  Count threshold used to binarize presence/absence.

- `min_n`  
  Minimum number of shared samples required for similarity scoring.

- `min_presence_score`  
  Threshold applied to composite presence similarity.

- `keep_unassigned`  
  If `TRUE`, retains DIRTY ASVs that fail assignment.

- `n_cores`, `parallel_method`  
  Control parallel execution (`"none"`, `"fork"`, `"cluster"`).

---

## Method Overview

AssignParents proceeds in five stages:

1. **Sequence identity filtering**  
   Computes DIRTY × CLEAN percent identity (Hamming or Levenshtein depending upon whether sequence length is/isn't equal across all ASVs). Sequence identity defines the set of eligible parent candidates.

2. **Presence composite scoring**  
   Combines prevalence metrics, including the phi coefficient, Positive Pointwise Mutual Information (PPMI), and F1 score. Each of these metrics were chosen since they measure complementary aspects of co-occurrence structure after read counts are binarized into presence and absence.

   The phi coefficient measures overall association strength between any two ASVs across the dataset, accounting for both joint presences and joint absences. PPMI measures positive co-occurrence compared to random expectation and and is insensitive to associations driven by ASVs with high background prevalence. The F1 score accounts for both precision and recall, measuring the degree with which candidate parent ASVs consistently co-occur with the DIRTY ASV without being overly permissive.

   Each metric is centered and scaled and then summed together. After summation, the composite score is again center-scaled to produce a composite that accurately captures ASV co-occurrence structure.

3. **Abundance composite scoring**  
   Combines abundance similarity using Spearman correlation and Hellinger distance. We used a composite of these two metrix because they measure complementary aspects of abundance: Spearman correlation measures monotonic agreement in rank abundance across all samples, while Hellinger distance measures compositional similarity that is invariable to differences in total abundance.

   Each abundance metric is centered and scaled with a mean of zero and S.D.of 1. These center-scaled metrics are then summed to form an intermediate abundance composite score, which is center-scaled again to produce a final composite abundance metric that weighs the values for the Spearman correlation and Hellinger distance equally.

4. **Presence and abundance composite**  
   Presence and abundance similarity metrics are summed together and center-scaled. This composite meta-score is used for downstream deterministic parent-child selections.

5. **Dominance filtering**  
   Removes biologically implausible parent-child assignments. A parent ASV cannot be more abundant than the child ASV.

6. **Deterministic parent selection**  
  Sequence identity is the primary determinant of parent–child relationships. Among ASVs that pass the minimum identity threshold, the composite score is used as a tiebreaker during deterministic parent selection when multiple candidate parents have comparable sequence similarity.

   A z-score–based neighborhood rule allows the algorithm to override small differences in sequence similarity. By default, Parent ASV candidates that fall within an identity neighborhood of 1 mismatch (`identity_neighborhood_mismatches = 1`) are considered near-equivalent. Within this neighborhood, if one lower-identity candidate has a composite meta-score that exceeds the best higher-identity candidate by at least 1.4 standard deviations (`neighborhood_z_margin = 1.4`), it will be selected as the parent even if it has slightly lower sequence similarity. This rule allows strong ecological support to reproducibly resolve near-equivalent identity matches in parent ASV selection.

Each parallelization method and the sequential method were tested and produce identical results.

---

## Compatibility

AssignParents is compatible with ASV-based workflows that produce:

- An ASV count matrix (ASVs × samples)
- A list of DNA sequences per ASV

This includes:

- DADA2
- Deblur
- UNOISE
- QIIME2-generated ASV feature tables (after export)
- Custom denoising pipelines

AssignParents does not require a specific upstream pipeline.

---

## Citation

If you use AssignParents in published work, please cite the conceptual framework described in:

Oppenheimer, P. et al. (2024). *ISME Communications*.  
https://academic.oup.com/ismecommun/article/5/1/ycaf124/8209421

AssignParents builds upon and extends the methodological principles described in that study.

A dedicated methodological manuscript describing the AssignParents algorithm and implementation is in preparation.

## License

MIT License
